The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
its flaky, but SAT is done.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SAT Solver in {{langRevEE}} }
\label{sec:prog}

%%%%%%%%%%%%%%%%%%%
% \subsection{SAT solver}
\label{sec:sat-solver}

The construction presented here is a novel SAT-solver that relies on
{{trace*}} for its solution. It varies from previous classical SAT
solvers in that there is not explicit search operation on the boolean
space. It also varies from previous quantum SAT solvers in that it
does uses a {{trace}} operation to compute a fixpoint which is the
solution following which an isomorphic clone operation allows us to
examine the result.

The construction here is not limited to boolean satisfiability, but
generalizes to any constraint satisfaction problem whose search space
can be represented by a recursive type and for which an isomorphic
clone operation can be constructed. That said, we will focus on the
limited case of boolean satisfiability here.

Consider the circuit below that has no observable output -- i.e. for
all inputs it is annihilating. This happens because it constructs a
boolean and its dual, negates one of them and attempts to collapse
them.
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}

We leverage the basic idea of the above circuit, to collapse all
values that are not solutions to the specified isomorphic constraint
{{f}}. The contraint {{f}} checks the inputs to see if they match a
constraint -- in the case of the SAT solver, {{f}} would be the
circuit expressing the boolean expression to be satisfied.  While the
output of {{f}} may contain many components, its must contain a
boolean indicating if the input satisfy {{f}}. The operation {{f}}
acts on some inputs and its adjoint, {{f^{dagger} }}, reconstructs the
inputs from the outputs.  The `trick' is the following: the special
output wire is used to `control' the negation of a boolean `control
wire'.  If the inputs do not satisfy {{f}}, the control wire is
negated.

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/sat1.pdf}
}
\end{center}

@@

The final detail we have to be concerned with is the handling of the
heap and garbage of {{f}}. Any boolean expression {{bool^n -> bool}}
can be compiled into the isomorphism {{f:bool^h*bool^n<->bool^g*bool}}
where the extra bits {{bool^h}} and {{bool^g}} are the heap and
garbage. Constructing such an isomorphism has been detailed before
\cite{Toffoli:1980,infeffects}.  However the computation of {{f}} is
meaningful {{iff}} the heap {{bool^h}} is initialized appropriately. We
can ensure that the heap has the appropriate initial value by checking
the heap and negating a second control wire, if the values do not
match.

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/sat2.pdf}
}
\end{center}

\roshan{Explain how the dotted part is implemented.}

Let us call this construction which maps inputs, heap and control
wires to input, heap and control wires as {{sat_f}}. The required
contraint solver is then constructed as shown:

@@

%%%%%%%%%%%%%%%%%%%
\subsection{GoI machine}

We can now encode the GoI machine of
Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. \amr{The machine uses bang
  which allows arbitrary duplication. Can we really do that???}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Advanced Example: SAT Solver in {{langRevEE}} }
\label{sec:prog}
\label{sec:sat-solver}

Additive traces correspond to iteration and multiplicative traces
correspond to fixpoint of constraints. Addtive traces have been
studied in the past \cite{infeffects} and here we focus on the
multiplicative trace.

The construction presented here is a novel SAT-solver that relies on
{{trace*}} for its solution. It varies from previous classical SAT
solvers in that there is not explicit search operation on the boolean
space. It also varies from previous quantum SAT solvers in that it
does uses a {{trace}} operation to compute a fixpoint which is the
solution following which an isomorphic clone operation allows us to
examine the result.  Though we focus on SAT here, this construction
can be generalized to any constraint satisfaction problem whose search
space can be represented by a recursive type and for which an
isomorphic clone operation can be constructed. Before we start, let us
recall two constructions:

\begin{enumerate}
\item Recall from previous work that for the types {{bool}} and, by
  extension, any tuple of {{bool}}, an isomorphic clone operation can
  be constructed (See Lemma 7.2 \cite{infeffects}).

\item Also, Sec. 3.3 \cite{infeffects} constructed the combinator
  {{if_{not} }}, which when generalized to two bits was
  {{toffoli}}. This construction maybe generalised to {{n}} bits as
  {{if^n_{not} }}.
\end{enumerate}

%% That said, we will focus on the limited case of boolean satisfiability
%% here.

\emph{A Total Annihilation Circuit.}  Consider the circuit below that
has no observable output -- i.e. for all inputs it is
annihilating. This happens because it constructs a boolean and its
dual, negates one of them and attempts to collapse them.

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}

\emph{Controlled Annihilation.}
We leverage the basic idea of the above circuit, to collapse all
values that are not solutions to the specified isomorphic constraint
{{f}}. The contraint {{f}} checks the inputs to see if they match a
constraint -- in the case of the SAT solver, {{f}} would be the
circuit expressing the boolean expression to be satisfied.  While the
output of {{f}} may contain many components, its must contain a
boolean indicating if the input satisfy {{f}}. The operation {{f}}
acts on some inputs and its adjoint, {{f^{dagger} }}, reconstructs the
inputs from the outputs.  The `trick' is the following: the special
output wire is used to `control' the negation of a boolean `control
wire'.  If the inputs do not satisfy {{f}}, the control wire is
negated.

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/sat1.pdf}
}
\end{center}

@@

\emph{Controlling the heap.}  The final detail we have to be concerned
with is the handling of the heap and garbage of {{f}}. Any boolean
expression {{bool^n -> bool}} can be compiled into the isomorphism
{{f:bool^h*bool^n<->bool^g*bool}} where the extra bits {{bool^h}} and
{{bool^g}} are the heap and garbage. Constructing such an isomorphism
has been detailed before \cite{Toffoli:1980,infeffects}.  However the
computation of {{f}} is meaningful {{iff}} the heap {{bool^h}} is
initialized appropriately. We can ensure that the heap has the
appropriate initial value by checking the heap and negating a second
control wire, if the values do not match -- i.e. the dotted part in
the diagram below and is {{if^h_{not} }} followed by a {{not}} on the
`heap control wire'.

\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/sat2.pdf}
}
\end{center}

% \roshan{Explain how the dotted part is implemented.}

\emph{Cloning the Solution.}.  Let us call the above construction
which maps inputs, heap and control wires to input, heap and control
wires as {{sat_f}}. The SAT-solver is is completed by tracing the
{{sat_f}} and cloning the `inputs'.

@@

When the solver is fed inputs initialized to {{true}}, the {{clone}}
will clone any inputs to {{sat_f}} that satisfies the constraints
imposed by {{f}} and the constraints on the {{heap}}. In the case of
unique-SAT the solver will produce exactly 1 or 0 solutions. In the
case of general SAT the solver will produce solutions as determined by
the top-level (Sec. \ref{sec:rat}).

%% %%%%%%%%%%%%%%%%%%%
%% \subsection{GoI machine}

%% We can now encode the GoI machine of
%% Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. \amr{The machine uses bang
%%   which allows arbitrary duplication. Can we really do that???}
